\documentclass[lnbip]{svmultln}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{ragged2e}
\usepackage[bookmarks=false,colorlinks,urlcolor=blue,%
linkcolor=magenta,citecolor=red,linktocpage=true,breaklinks=true]{hyperref}
\usepackage{cleveref}
\usepackage{xspace}
\usepackage{graphics}
\usepackage{inconsolata}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{listings}
\lstset{escapeinside={<@}{@>}}
\usepackage{xparse}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{pgf-umlsd}

% \linespread{.97}
% \usepackage[backend=bibtex,style=ieee,giveninits,maxnames=2,minnames=1]
% {biblatex}

% \AtEveryBibitem{\clearfield{pages}}
% \AtEveryBibitem{\clearfield{volume}}
% \AtEveryBibitem{\clearfield{number}}
% \addbibresource{biblio.bib}

\include{macros}

\usepackage[textsize=small,linecolor=teal,%
bordercolor=teal,backgroundcolor=lime!25]{todonotes} % for margin

\begin{document}

\mainmatter              % start of the contribution

\title{Linguistic Abstractions for\\ Interoperability of IoT Platforms}

\titlerunning{Language Abstractions for Interoperability of IoT
 Platforms}  % abbreviated title (for running head)

 \author{Maurizio Gabbrielli\inst{1} \and
 Saverio Giallorenzo\inst{2} \and
 Ivan Lanese\inst{1} \and Stefano Pio
 Zingaro \inst{1}}
%
\authorrunning{Gabbrielli M. et al.}   % abbreviated author list (for
% running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Maurizio Gabbrielli, Saverio Giallorenzo, Ivan Lanese, Stefano Pio
Zingaro}
%
\institute{Universit\`a di Bologna / INRIA \\
\and
University of Southern Denmark}

\maketitle

% The Internet of Things (IoT) promotes the communication among
% heterogeneous entities, from small sensors to Cloud systems. However,
% this is realized using a wide range of communication media and data
% protocols, usually incompatible with each other. Thus, IoT systems
% tend to grow as homogeneous isolated platforms, which hardly
% interact. In this paper we propose a language-based approach for
% communication technology integration to achieve a higher degree of
% interoperability among disparate IoT platforms. We build on the Jolie
% programming language, which provides uniform linguistic abstractions
% to exploit heterogeneous communication stacks, allowing the programmer
% to specify in a declarative way the desired stack, and to easily
% change it, even at runtime.  Jolie currently supports the main
% technologies from Service-Oriented Computing, such as TCP/IP,
% Bluetooth, and RMI at transport level, and HTTP and SOAP at
% application level. We integrate in Jolie the two most adopted
% protocols for IoT communication, i.e., CoAP and MQTT. We report our
% experience, including a case study in the home automation area, and we
% present high-level concepts valuable both for the general
% implementation of interoperable systems and for the development of
% other language-based solutions.

\begin{abstract}
The Internet of Things (IoT) advocates for multi-layered platforms---from edge
devices to Cloud nodes---where each layer adopts its own communication
standards (media and data formats). While this freedom is optimal for in-layer
communication, it puzzles cross-layer integration due to incompatibilities
among standards. Also enforcing a unique communication stack within the same
IoT platform is not a solution, as it leads to the current phenomenon of ``IoT
islands'', where disparate platforms hardly interact with each other. In this
paper we tackle the problem of IoT cross-layer and cross-platform integration
following a language-based approach. We build on the Jolie programming
language, which provides uniform linguistic abstractions to exploit
heterogeneous communication stacks, allowing the programmer to specify in a
declarative way the desired stack, and to easily change it, even at runtime.
Jolie currently supports the main technologies from Service-Oriented Computing,
such as TCP/IP, Bluetooth, and RMI at transport level, and HTTP and SOAP at
application level. We integrate in Jolie the two most adopted protocols for IoT
communication, i.e., CoAP and MQTT. We report our experience on a case study on
Cloud-based home automation, and we present high-level concepts valuable both
for the general implementation of interoperable systems and for the development
of other language-based solutions.
\end{abstract}

\section{Introduction}
\label{sec:intro}
\input{intro}

\section{JIoT: Jolie for IoT}
\label{sec:challenges}
\input{challenges}

\section{Supporting Constrained Application Protocol in Jolie}
\label{sub:coap}
\input{coap}

\section{Supporting Message Queue Telemetry Transport in Jolie}
\label{sub:mqtt}
\input{mqtt}

\section{Implementation}
\label{sub:implementation}
\input{implementation}

\section{Case Study}
\label{sub:case}
\input{case}

\section{Related Work}
\label{sec:related}
\input{related}

\section{Discussion and Conclusion}
\label{conclusion}
\input{conclusion}

\subsubsection*{Acknowledgments} % (fold)
We thank Marco Di Felice, Luca Bedogni, and Federico Montori for useful
suggestions and comments.

% section section_name (end)

% \vspace{5em}

\bibliographystyle{ieeetr}
{\bibliography{biblio}}

\end{document}
